Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-03-15 young_pp_discs.tex
minor tweaks

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (figure \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}) it can be used to view rotation of the central part of the disc without risk of contamination from emission from shocked material along the outflow. The OCS lines are unique amongst the lines simulated in that it is not visible throughout all the regions in the disc where it is found, the line is only seen at radii <16$\,au$ whereas the molecules have a significant abundance out to $\sim$30$\,$au. This is because the OCS line, which has the highest upper energy level among the selected transitions, traces only the hottest and densest regions of the disc where the high energy states are populated. As a result, the innermost spiral structure is best seen in OCS (26 - 25), as shown in figure \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, because of the presence of the protostar and other hot material within 2$\,$au of the protostar, not considered in our simple model, we expect absorption to also be observed toward the central region of the disc.\smallskip

@@

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate their observation with the completed ALMA in the 26th most extended configuration (out of a total of 28), with the longest baseline of 14.4km and giving beam sizes between 0.02 and 0.03 arcseconds at the selected frequencies, giving a beam size close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
  \centering
  \begin{minipage}{80mm}
    \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
    \label{sigmas}
    \begin{tabular}{c||c|c|c|c}
      \hline
      Species & Transition & Frequency & Upper energy level & Sensitivity\\
      \hline
      C$^{17}$O & 3$\rightarrow$2 & 337.06GHz & 32.3$\,$K & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
      OCS & 26$\rightarrow$25 & 316.15GHz & 25.7$\,$K & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
      HCO$^+$ & 3$\rightarrow$2 & 267.56GHz & 205$\,$K & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
      H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62GHz & 34.9$\,$K & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
      \hline
    \end{tabular}
  \end{minipage}
\end{table*}
In order to clean the images which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in Table \ref{sigmas}.\smallskip

@@

 \caption{CASA simulation of the C17O 3$\rightarrow$2 line with the disc inclined at 15$^\circ$ to edge on {\bf Top:} integrated intensity map. {\bf bottom:} position velocity diagram. In the position velocity diagram the spiral structures can be seen as narrow finger-like structures extending from the origin.}
 \label{15deg}
\end{figure}

@@

These results demonstrate that with next generation millimetre interferometers we will be able to probe the very earliest stages of circumstellar discs which feed their growing protostar and eventually could become a planetary system. Our results show that these types of observations would be able to detect whether or not these young disc are gravitationally unstable. If such gravitationally instabilities exist they would have significant implications on the evolution of the disc, such as fragmentation and planetesimal formation \citep{Boley2009,Johnson2013,Gibbons2012} and chemical evolution I2011.

>>>>

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (figure \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}) it can be used to view rotation of the central part of the disc without risk of contamination from emission from shocked material along the outflow. The OCS lines are unique amongst the lines simulated in that it is not visible throughout all the regions in the disc where it is found, the line is only seen at radii $<$16$\,$au whereas the molecules have a significant abundance out to $\sim$30$\,$au. This is because the OCS line, which has the highest upper energy level among the selected transitions, traces only the hottest and densest regions of the disc where the high energy states are populated. As a result, the innermost spiral structure is best seen in OCS (26 - 25), as shown in figure \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, because of the presence of the protostar and other hot material within 2$\,$au of the protostar, not considered in our simple model, we expect absorption to also be observed toward the central region of the disc.\smallskip

@@

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate their observation with the completed ALMA in the 26th most extended configuration (out of a total of 28), with the longest baseline of 14.4km and giving beam sizes between 0.02 and 0.03 arcseconds at the selected frequencies. This gives a beam size close to the size of the spiral structure in the disc, allowing for the best sensitivity whilst being able to resolve the spiral structure. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
  \centering
  \begin{minipage}{80mm}
    \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
    \label{sigmas}
    \begin{tabular}{c||c|c|c|c}
      \hline
      Species & Transition & Frequency & Upper energy level & Sensitivity\\
      \hline
      C$^{17}$O & 3$\rightarrow$2 & 337.06GHz & 32.3$\,$K & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
      OCS & 26$\rightarrow$25 & 316.15GHz & 25.7$\,$K & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
      HCO$^+$ & 3$\rightarrow$2 & 267.56GHz & 205$\,$K & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
      H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62GHz & 34.9$\,$K & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
      \hline
    \end{tabular}
  \end{minipage}
\end{table*}
In order to clean the images which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in Table \ref{sigmas}.\smallskip

@@

 \caption{CASA simulation of the C17O 3$\rightarrow$2 line with the disc inclined at 15$^\circ$ to edge on. {\bf Left:} Integrated intensity map. {\bf Right:} Position velocity diagram. In the position velocity diagram the spiral structures can be seen as narrow finger-like structures extending from the origin.}
 \label{15deg}
\end{figure}

@@

These results demonstrate that with next generation millimetre interferometers we will be able to probe the very earliest stages of circumstellar discs which feed their growing protostar and eventually could become a planetary system. Our results show that these types of observations would be able to detect whether or not these young disc are gravitationally unstable. If such gravitationally instabilities exist they would have significant implications on the evolution of the disc, such as fragmentation and planetesimal formation \citep{Boley2009,Johnson2013,Gibbons2012} and chemical evolution in the disc (I2011).
